# Single continuous explanatory {#bino-glm-single-cont}

## Introduction to the example {#intro-6}
The erythrocyte sedimentation rate (ESR) is a routine test for inflammation where a reading greater than 20mm/hr indicates inflammation. However, fibrinogen levels also affect ESR. The aim of this analysis is to determine whether an ESR greater than 20mm/hr can be predicted from fibrinogen levels.
The data in [plasma.txt](data-raw/plasma.txt) are a simplified version of Collett and Jemain's (1985) data which comes with the **`HSAUR`** package [@HSAUR]. There are two variables: ESR where 0 indicates less than or equal to 20mm/hr and 1 indicates greater than 20mm/hr; and fibrinogen in arbitrary units.


```{r echo=FALSE}
plasma <- read_table2("data-raw/plasma.txt")
plasma %>%
  knitr::kable() %>%
  kable_styling()  %>%
  scroll_box(height = "300px")
```


:::key
The response is ESR where 0 indicates less than or equal to 20mm/hr and 1 indicates greater than 20mm/hr. Fibrinogen is the explanatory variable.
:::


We will import the data with the `read_table2()` function and plot it with `ggplot()`.

```{r eval=FALSE}
plasma <- read_table2("data-raw/plasma.txt")
```


```{r}
# a default scatter plot of the data
ggplot(data = plasma, aes(x = fibrinogen, y = ESR)) +
  geom_point()
```
Those with lower ESR do seem to have lower fibrinogen and half of those with high ESR have high fibrinogen. However, we have relatively few high ESR observations. Is the association significant?

## Applying and interpreting `glm()` {#apply-interp-6}

We build a generalised linear model of ESR explained by fibrinogen with the `glm()` function as follows:

```{r}
mod <- glm(data = plasma, ESR ~ fibrinogen, family = binomial)
```
Printing `mod` to the console gives us the estimated model parameters:
```{r}
mod
```


```{r echo=FALSE}
res <- summary(mod)
b0 <- res$coefficients["(Intercept)", "Estimate"]
b1 <- res$coefficients["fibrinogen", "Estimate"]
b0se <- res$coefficients["(Intercept)", "Std. Error"]
b1se <- res$coefficients["fibrinogen", "Std. Error"]
zval <- res$coefficients["fibrinogen", "z value"]
b0_resp <- exp(b0)
b1_resp <- exp(b1)
b1se_resp <- exp(b1se)
resid_dev <- res$deviance
df_resid <- res$df.residual
null_dev <- res$null.deviance
df_null <- res$df.null
mod_dev <- null_dev - resid_dev
AIC <- res$aic
df <- res$df[2]

if (res$coefficients["(Intercept)", "Pr(>|z|)"] < 0.001) {
        b0p = "< 0.001"
        }
if (res$coefficients["(Intercept)", "Pr(>|z|)"] > 0.001) {
        b0p = paste("=", round(res$coefficients["(Intercept)", "Pr(>|z|)"], 3))
        }
if (res$coefficients["fibrinogen", "Pr(>|z|)"] < 0.001) {
        b1p = "< 0.001"
        }
if (res$coefficients["fibrinogen", "Pr(>|z|)"] > 0.001) {
        b1p = paste("=", round(res$coefficients["fibrinogen", "Pr(>|z|)"], 3))
}

res2 <- anova(mod, test = "Chisq")
if (res2$`Pr(>Chi)`[2] < 0.001) {
        modelp = "< 0.001"
        }
if (res2$`Pr(>Chi)`[2] > 0.001) {
        modelp = paste("=", round(res2$`Pr(>Chi)`[2], 3))
}

```

We will postpone discussing the information in the last three lines until we view the model summary.

$\beta_{0}$ is labelled "(Intercept)" and $\beta_{1}$ is labelled "fibrinogen". Thus the equation of the line is:

<center> $ln\left(\frac{P(ESR >20)}{1-P(ESR >20)}\right)$ = `r b0` $+$ `r b1`$\times fibrinogen$ </center>

The fact that the estimate for fibrinogen (`r b1`) is positive tells us the probability of having an ESR greater than 20 mm/hr increases as fibrinogen increases.

The fact that the estimate for the intercept (`r b0`) is negative tells us the probability of having an ESR greater than 20 mm/hr is lower than the probability of having an an ESR less than 20 mm/hr at a fibrinogen of zero.

Notice that we are thinking of having an ESR greater than 20 mm/hr as success. Which of the two outcomes you consider success, or 1, is arbitrary. You just need to make sure the outcome you are classing as success is coded as 1 in your data set. It need not be considered a success biologically. In this example, it means the probability we are modelling is the probability of having an ESR greater than 20 mm/hr.

These estimates are on the scale of the link function - they are log odds.

To understand the parameters the on the scale of the response we apply the inverse of the link function. We need to first exponentiate them, then interpret them as odds.

```{r}
exp(mod$coefficients)
```

`r b0_resp` is the odds of an ESR >20 at a fibrinogen of 0. Odds are P(ESR > 20) / P(ESR < 20). As this is less than 1, the probability of an ESR >20 must be smaller than the probability ESR < 20 at a fibrinogen of 0. 

`r b1_resp` is the factor by which this odds changes with each unit of fibrinogen. As this is greater than 1, it's going up, not down.

An alternative way to write the equation of the line is:

<center> $\frac{P(ESR >20)}{P(ESR <20)}$ = `r b0_resp` $\times$ `r b1_resp`$^{fibrinogen}$ </center>

:::key

:::

More information including statistical tests of the model and its parameters is obtained by using `summary()`:

```{r}
summary(mod) 
```
The `Coefficients` table gives the estimated $\beta_{0}$ and $\beta_{1}$ again but along with their standard errors and tests of whether the estimates differ from zero. The estimated value for the intercept is `r b0` $\pm$ `r b0se` and this differs significantly from zero ($p$ `r b0p`). The estimated value for the slope is `r b1` $\pm$ `r b1se`, also differs significantly from zero ($p$ `r b1p`).

Towards the bottom of the output there is information about the model fit. The null deviance (what exists if we predict the odds of ESR > 20 from an intercept, $\beta_{0}$, only model) is `r null_dev` with `r df_null` degrees of freedom and the residual deviance (left over after our model is fitted) is `r resid_dev` with `r df_resid` $d.f.$. The model fits a single parameter (after $\beta_{0}$), $\beta_{1}$, and thus accounts for 1 $d.f.$ for a reduction in deviance by `r mod_dev`.

To get a test of whether the reduction in deviance is significant for each term in the the model formula we use:

```{r}
anova(mod, test = "Chisq")
```

There is a significant reduction in deviance for our model (p `r modelp`).

## Getting predictions from the model {#get-pred-6}

The `predict()` function returns the predicted values of the response. To add a column of predicted values to the dataframe:


```{r}
# plasma$pred <- predict(mod, type = "response")
```
We need to specify they should be on the scale of the responses (i.e., probabilities), not on the link function scale (log odds).

This gives predictions for the actual $x$ values used. If you want predictions for other values of $x$ you need to creating a data frame of the $x$ values from which you want to predict

For example, to predict for fibrinogen from 0 to 6 km in steps of 1 km:
```{r}
predict_for <- data.frame(fibrinogen = seq(0, 6, 1))
```


```{r}
predict_for$pred <- predict(mod, newdata = predict_for, type = "response")
```

```{r echo=FALSE}
pred_0 <- predict_for$pred[1]
pred_1 <- predict_for$pred[2]
pred_2 <- predict_for$pred[3]
pred_6 <- predict_for$pred[7]
```
## Using the predictions to understand the odds

You can use the predictions at fibrinogen levels of 0 and 1 to help you understand odds and the model coefficients.

*a bit more to come, soon, honest!*
<!-- An odds is the probability of something happening divided by the probability of it not happening.  -->
<!-- # At a grain size of 0, P(present) is 0.1614302, therefore p(absent) is 1-0.1614302 -->
<!-- # 0.1614302/(1-0.1614302) = 0.1925066. Look! that's exp(b0)! -->
<!-- # How does this change from grain size of 0 to grain size of 1? -->
<!-- # At a grain size of 1, P(present) is 0.9699368, therefore p(absent) is 1-0.9699368 -->
<!-- # the factor by which the odds change is the odds of present at grainsize 1 / -->
<!-- # odds of present at grainsize 0 -->
<!-- # (0.9699368/(1-0.9699368)) / (0.1614302/(1-0.1614302)) -->
<!-- # = 167.5956 Look, that's exp(b1) -->
<!-- # Odds are not intuitive for most people. I almost always use the predict function to calculate probabilities to help me understand the effects in my model. -->
<!-- # You can still make quick judgement on the basis of the coefficients -->
<!-- # negative b0 means probability of absence (or died/0/no) is higher than probability of presence (or survived/1/yes); positive b0 means probability of presence is higher than probability of absence -->
<!-- # negative b0 means odds of get lower, presence gets less likely -->
<!-- # positive b0 means odds of get higher, presence gets less likely -->


## Creating a figure {#fig-6}

```{r fig-plasma}
ggplot(data = plasma, aes(x = fibrinogen, y = ESR)) +
  geom_point() +
geom_smooth(method = "glm",
              method.args = list(family = "binomial"),
              se = FALSE,
            colour = "black") +
  scale_x_continuous(expand = c(0, 0),
                     limits = c(0, 6),
                     name = "Fibrinogen (units)") +
    scale_y_continuous(expand = c(0, 0.03),
                      breaks = seq(0, 1, 0.1),
                     limits = c(0, 1.05),
                     name = "Probability of ESR >2 0") +
  theme_classic()
  
```


## Reporting the results {#report-6}


The odds of erythrocyte sedimentation rate (ESR) greater than 20 mm/hr significantly increases with increasing fibrinogen levels by a factor of `r b1_resp` for each unit of fibrinogen (p `r modelp`). See Figure \@ref(fig:fig-plasma-report). As an example, the probability of having an ESR greater than 20 mm/hr at 2 units `r pred_2` is which rises to `r pred_6` at 6 units.

(ref:fig-plasma-report) Incidence erythrocyte sedimentation rate (ESR) greater than 20 mm/hr with fibrinogen levels. The line gives predictions for a GLM with binomial distributed errors, $\frac{P(ESR >20)}{P(ESR <20)}$ = `r b0_resp` $\times$ `r b1_resp`$^{fibrinogen}$.

```{r fig-plasma-report, ref.label = 'fig-plasma', echo = FALSE, fig.height=4, fig.width=4, fig.cap="(ref:fig-plasma-report)", out.width="60%"}
```
