# (PART) USING `glm()` FOR POISSON RESPONSES {-}

# Single continuous predictor 

## Introduction to the example

Does proximity to a nuclear power plant alter the risk of cancer?
Number of cases of cancer reported by a clinic by its distance from nuclear plant (note not good epidemiology). each row is a clinic
[caes.txt](data-raw/cases.txt)


```{r}
cases <- read_table2("data-raw/cases.txt")
```


```{r}
ggplot(data = cases, aes(x = distance, y = cancers)) +
  geom_point()
```

## Applying and interpreting lm()

```{r}
mod <- glm(data = cases, cancers ~ distance, family = poisson)


```

```{r}
mod
```
```{r}
mod$coefficients
```


```{r echo=FALSE}
b0 <- mod$coefficients["(Intercept)"]
b1 <- mod$coefficients["distance"]
b0_resp <- exp(b0)
b1_resp <- exp(b1)
```

The fact that the estimate for distance (`r b0`) is negative tells us that as distance increases, the number of cancers reported goes down.

These estimates are on the scale of the link function, that is, they are logged (to the base e, natural logs) in this case.

To understand thm on the scale of the response we apply the inverse of the ln function, the exp() function

```{r}
exp(mod$coefficients)
```

$\beta_{0}$ is labelled "(Intercept)" and $\beta_{1}$ is labelled "distance". Thus the equation of the line is:

<center> $cancers$ = `r b0` $\times$ `r b1`$^{distance}$ </center>

The model predicts there will be `r b0_resp` cancers at a clinic at no distance from the power plant.

Recall that for a linear model with one predictor, the second estimate is the amount added to the intercept when the predictor changes by one value. Since this is glm with a log link, the value of `r b1_resp` is amount the intercept is multiplied by. Thus the model predicts there will be `r b0_resp` $\times$ `r b1_resp` =  `r b0_resp*b1_resp` cancers 1 km away and `r b0_resp` $\times$ `r b1_resp` $\times$ `r b1_resp` =  `r b0_resp*b1_resp*b1_resp` cancers 2 km away. That is: $\beta_{0}$ $\times$ $\beta_{0}^n$ mm at $n$ km away.




```{r}
summary(mod) 
```


To get a test of the model overall

```{r}
anova(mod, test = "Chisq") 

```
```{r}
pchisq(22.732, 1, lower.tail = F)
```

## Getting predictions from the model

The `predict()` function returns the predicted values of the response. To add a column of predicted values to the dataframe:
we need to specify they should be on the scale of the responses, no the link function scale.

```{r}
cases$pred <- predict(mod, type = "response")
```




This gives predictions for the actual x values used. If you want predicts for other values of x you need to creating a data frame of the x values from which you want to predict


```{r}
predictions <- data.frame(distance = seq(0, 180, 10))
```


```{r}
predictions$pred <- predict(mod, newdata = predictions, type = "response")
```

