# Two explanatory variables {#pois-glm-two-cont}

## Introduction to the example

The number of cases of cancer reported by a clinic and its distance, in kilometres, from a nuclear plant were recorded and the data are in [cases.txt](data-raw/cases.txt). Researchers wanted to know if proximity to the nuclear power plant influenced the incidence of cancer. Bear in mind this is not great epidemiology - there would be very many other factors influencing the occurrence and reporting of cancer cases at a clinic.


```{r echo=FALSE}
cases <- read_table2("data-raw/cases.txt")
cases %>% 
  knitr::kable() %>%
  kable_styling()  %>%
  scroll_box(width = "300px", height = "300px")
```


:::key
There are 2 variables: the response, `cancers`, is the number of cancer cases reported at a clinic and `distance`, gives the clinic's distance from the nuclear plant. 
:::


We will import the data with the `read_table2()` function and plot it with `ggplot()`.

```{r eval=FALSE}
cases <- read_table2("data-raw/cases.txt")
```


```{r}
# a default scatter plot of the data
ggplot(data = cases, aes(x = distance, y = cancers)) +
  geom_point()
```
Most of the clinics reporting no cases seem to be more distance from the nuclear plant and those reporting the highest numbers are within 50km.

## Applying and interpreting `glm()`

We build a generalised linear model of the number of cases explained by the distance with the `glm()` function as follows:

```{r}
mod <- glm(data = cases, cancers ~ distance, family = poisson)
```
Printing `mod` to the console gives us the estimated model parameters:
```{r}
mod
```


```{r echo=FALSE}
res <- summary(mod)
b0 <- res$coefficients["(Intercept)", "Estimate"]
b1 <- res$coefficients["distance", "Estimate"]
b0se <- res$coefficients["(Intercept)", "Std. Error"]
b1se <- res$coefficients["distance", "Std. Error"]
zval <- res$coefficients["distance", "z value"]
b0_resp <- exp(b0)
b1_resp <- exp(b1)
resid_dev <- res$deviance
df_resid <- res$df.residual
null_dev <- res$null.deviance
df_null <- res$df.null
mod_dev <- null_dev - resid_dev
AIC <- res$aic
df <- res$df[2]

if (res$coefficients["(Intercept)", "Pr(>|z|)"] < 0.001) {
        b0p = "< 0.001"
        }
if (res$coefficients["(Intercept)", "Pr(>|z|)"] > 0.001) {
        b0p = paste("=", round(res$coefficients["(Intercept)", "Pr(>|t|)"], 3))
        }
if (res$coefficients["distance", "Pr(>|z|)"] < 0.001) {
        b1p = "< 0.001"
        }
if (res$coefficients["distance", "Pr(>|z|)"] > 0.001) {
        b1p = paste("=", round(res$coefficients["distance", "Pr(>|z|)"], 3))
        }
```
$\beta_{0}$ is labelled "(Intercept)" and $\beta_{1}$ is labelled "distance". Thus the equation of the line is:

<center> $ln(cancers)$ = `r b0` $\times$ `r b1`$\times distance$ </center>

The fact that the estimate for distance (`r b1`) is negative tells us that as distance increases, the number of cancers reported goes down.

These estimates are on the scale of the link function, that is, they are logged (to the base e, natural logs) in this case.

To understand the parameters the on the scale of the response we apply the inverse of the $ln$ function, the `exp()` function

```{r}
exp(mod$coefficients)
```

So
<center> $cancers$ = `r b0_resp` $\times$ `r b1_resp`$^{distance}$ </center>

The model predicts there will be `r b0_resp` cancers at a clinic at no distance from the power plant.

Recall that for a linear model with one predictor, the second estimate is the amount added to the intercept when the predictor changes by one value. Since this is glm with a log link, the value of `r b1_resp` is amount the intercept is multiplied by for each unit increase of distance. Thus the model predicts there will be `r b0_resp` $\times$ `r b1_resp` =  `r b0_resp*b1_resp` cancers 1 km away and `r b0_resp` $\times$ `r b1_resp` $\times$ `r b1_resp` =  `r b0_resp*b1_resp*b1_resp` cancers 2 km away. That is: $\beta_{0}$ $\times$ $\beta_{0}^n$ mm at $n$ km away.

You can work these out either by exponentiating the coefficients and then multiplying the results or by adding the coefficients and exponentiating

Exponentiate coefficients then multiply:
```{r}
# 1km away
exp(b0) * exp(b1)

# 2km away
exp(b0) * exp(b1) * exp(b1)

# 10km away
exp(b0) * exp(b1)^10

```
Add the coefficients then exponentiate the sum: 
```{r}
# 1km away
exp(b0 + b1)

# 2km away
exp(b0 + b1 + b1)

# 10km away
exp(b0 + 10*b1)
```
Usually, we use the `predict()` function to make predictions for particular distances.

More information including statistical tests of the model and its parameters is obtained by using summary():

```{r}
summary(mod) 
```
